Reduced mitochondrial transcription sensitizes acute myeloid leukemia cells to BCL-2 inhibition

Version 1

Abstract

Abstract

Overcoming drug-resistance and the subsequent relapse that often occurs with monotherapy is crucial in the treatment of acute myeloid leukemia. We here demonstrate that therapy-resistant leukemia initiating cells can be targeted using a novel inhibitor of mitochondrial transcription (IMT). The compound inhibits mitochondrial RNA polymerase activity and sensitizes the resistant population to the induction of apoptosis. In vitro studies on acute myeloid leukemia cells demonstrate that IMT prevents cell proliferation, and together with a selective BCL-2 inhibitor, venetoclax, induces apoptosis and suppress oxidative phosphorylation (OXPHOS) synergistically. AML mouse models treated with IMT in combination with venetoclax show prolonged survival in venetoclax-resistant models. Our findings suggest that certain therapy-resistant leukemia cell populations display a unique dependency on mitochondrial transcription and can be targeted with IMT.

Introduction

Acute myeloid leukemia (AML) comprises a devastating and heterogeneous group of diseases, for which patient survival rates are dismal despite several new treatment options. Current cares are still mainly based on unselective chemotherapy but for most patients also with addition of more directed treatments after patient stratification based on genetic aberrations found at diagnosis. However, these strategies lead to cures in only a fraction of AML patients due to therapy resistance developed by the leukemic cells leading to relapse. Curative treatment usually needs allogenic stem cell transplantation, which is not applicable in many cases, highlighting the critical need to develop different approaches. The anti-apoptotic molecule BCL-2 has been a therapeutic target in lymphoid and myeloid leukemia, as in many cancers overexpression of BCL-2 results in dysregulation of apoptosis (1). ABT-199 (venetoclax), is a specific BH3-mimetic that causes apoptosis by preventing BCL-2 from binding pro-apoptotic BAX and BAK1 proteins (2). However, complete responses occur in just ∼20% of AML patients treated with Venetoclax (3). In addition, many patients relapse, with regrowth initiated by therapy-resistant leukemic clones, suggesting that single agent administration is unlikely to yield lasting responses in most cases. Venetoclax is therefore used in combination with conventional therapies. Together with hypomethylating agents, Venetoclax is now standard of care for patients deemed unfit for intensive induction therapy (4).

Resistance to venetoclax may result from overexpression of the anti-apoptotic proteins MCL1 or BCL2L1 (5), or through changes in levels of other anti-apoptotic molecules (6). Aberrant pro-survival regulation can also be achieved by inactivation of TP53 function, which controls many aspects of apoptosis (6, 7). MCL1-dependent resistance to venetoclax can be overcome when the compound is combined with the MDM2 inhibitor RG7388, which causes p53 activation. A combination of BCL-2 inhibition and TP53 activation may thus provide an interesting therapeutic approach for treatment of AML (8).

Other combination approaches that promote apoptosis may also be of interest. The majority of ATP required for leukemia stem cell (LSC) function is generated by oxidative phosphorylation (OXPHOS) (9, 10). Interestingly, venetoclax-resistant leukemic cells display increased levels of OXPHOS and are protected from mitochondrial stress (7). These data suggest approaches to specifically target OXPHOS in leukemia may be an attractive option. In support of this notion, inhibition of mitochondrial protein translation by antibiotics that reduce OXPHOS capacity results in a selective cytotoxic effect on leukemic cell when used in combination with tyrosine kinase inhibitors (11). Furthermore, LSCs from de novo AML patients display a unique dependency on amino acid metabolism to fuel OXPHOS. When venetoclax is combined with the hypomethylating agent azacytidine, the two compounds act in concert to inhibit amino acid uptake, leading to a drop in OXPHOS and thus cell death (12). More recently, AML therapy using a combination of cytarabine (AraC) and venetoclax has highlighted the central role of mitochondrial adaptation during treatment (10, 13), indicating the need for a mitochondrial directed combination therapy for AML. Such a therapy may help to overcome existing treatment resistance.

Recently we developed novel, first-in-class inhibitors of mitochondrial transcription (IMTs) (14). IMT treatment specifically targets mitochondrial RNA polymerase (POLRMT) and reduce transcription of mitochondrial DNA, which codes for thirteen key components of the OXPHOS machinery, including subunits of respiratory complexes I, III, IV, and the ATP synthase. As a consequence, IMT treatment impairs OXPHOS system biogenesis in a dose-dependent manner and decreases cell viability in various tumor cells. Surprisingly, several weeks of per oral IMT treatment is well tolerated in mice and does not cause toxicity in normal tissues, despite inducing a strong anti-tumor response in human cancer xenografts (14).

Here we examine AML cell lines for their sensitivity for treatment with IMT. We find that a combination of IMT and venetoclax displays a synergistic on cell growth and induction of apoptosis. We demonstrated the efficacy of this approach in leukemia mouse models generated by human AML cell lines as well as primary samples from AML patients.

Results

Synergistic effect of IMT in combination with venetoclax on AML cell lines

We were curious whether our inhibitors of mitochondrial transcription (IMTs) (14) could be effective against AML cells. The coumarin class of IMTs was optimized by medicinal chemistry to generate lead IMT compounds with improved metabolic stability in mice, such as LDC204857, herein simply referred to as IMT (Fig 1A, Fig. S1 and Table S1). IMT is a close analog of previously described IMT1B (Fig. S1) with a highly similar biochemical and cellular inhibition profile (14). It shows an extended half-life in vivo due to its increased metabolic stability (Table S1). Thus, by using IMT we could consistently reduce the mouse dose to 50 mg/kg per oral once daily, which is very well tolerated and does not exert any adverse effects in mice.

IMT+venetoclax combination treatment synergistically inhibits cell viability/growth of AML cell lines in vitro.

(A) Chemical structure of IMT (LDC204857) (B) AML cell lines were treated with single agents IMT or venetoclax at a range of concentrations (0.1–1000 nM and 0.01–100 nM, respectively), and assessed by cell viability tests after three and six days. (C) Synergism analysis of AML cell lines treated with a combination of IMT (0.05, 0.1, 0.2 μM for MV4-11 and MOLM-13, and 0.1, 0.5, and 1 μM for THP-1 and KG-1) and venetoclax (Vex; 1, 2, 4 nM for MV4-11 and MOLM-13, and 1, 5, 10 nM for THP-1 and KG-1) for six days. ‘S’ represents synergy of the combination treatment, evaluated using the Bliss independent model. Proliferation percentages are normalized to DMSO treated. All results are from three independent experiments.

Using four AML cell lines with different expression of TP53 (MV4-11 and MOLM-13 expressing wild type TP53; THP-1 and KG-1 expressing mutated TP53), we monitored the effects of increasing concentrations of IMT on cellular growth using a luminescence-based cell viability assay. Only marginal effects were observed after a 72-h period of exposure to IMT, but all cell lines displayed dose-dependent antileukemic activity after 144 h of treatment (Fig. 1B), in line with our previous observations (14). We also monitored venetoclax sensitivity by assessing cell viability, and noted that, in line with published work (16), a dose-dependent reduction in cell growth occurs after three days and becomes more pronounced after 6 days (Fig. 1B).

To then investigate whether IMTs show a synergistic effect when combined with venetoclax, we conducted a combination study in AML cell lines (Fig. 1C). After six days of treatment, we found a clear synergistic effect on cell proliferation in MV4-11 and MOLM-13 cell lines, which express wild type TP53, as determined by combination analysis (Bliss-independence model (15)). Treatment of THP-1 and KG-1 cell lines, which harbor mutated TP53, only revealed a synergistic effect when using higher concentrations of both IMT and venetoclax (Fig. 1C, additional treatment data are shown in Fig. S2).

Combination of IMT and venetoclax synergistically induces apoptosis and loss of mitochondrial membrane potential in MV4-11 and MOLM-13 cells

To investigate the factors involved in the inhibition of cell growth with IMT and venetoclax, AML cell lines were treated with single agents (IMT or venetoclax) or a combination of the two (Fig. 2A), and the stages of apoptosis were monitored using Annexin V and PI staining. No detectable increase in apoptosis was induced by IMT exposure alone in any AML cell line. Venetoclax treatment resulted in both early and late apoptosis induction in MV4-11 and MOLM-13 cell lines, and employing a combination of IMT and venetoclax significantly enhanced apoptosis induction in these wild type cell lines. On the other hand, combination treatment did not induce apoptosis in THP-1 and KG-1 mutated cell lines (Fig. 2A).

Combination treatment significantly enhances apoptosis induction in AML cell lines.

AML cell lines were treated with IMT (500 nM) and venetoclax (5 nM) for 72–96 h. (A) Apoptosis was determined by Annexin V/PI staining of the cells and analyzed by flow cytometry. (B) Loss of mitochondrial membrane potential was calculated as the percentage of depolarized (apoptotic) cells out of the total cell population. Error bars represent standard error of the mean where * represents P < 0.05, **P < 0.01, and ***p < 0.001.

One distinctive feature of apoptosis is the disruption of mitochondrial function, which can be measured as loss of mitochondrial membrane potential due to membrane depolarization. As a combination of IMT and venetoclax could synergistically induce apoptosis, we were curious whether combination drug treatment thus affects mitochondrial membrane potential. We observed that venetoclax-sensitive cell lines (MV4-11 and MOLM-13) showed a dramatic loss of mitochondrial membrane potential when treated with both venetoclax and IMT (Fig. 2B). On the other hand, mitochondrial membrane potential in THP-1 cells was not affected by single nor by combination treatment. Interestingly, although KG-1 was venetoclax insensitive, the combination treatment significantly enhanced loss of mitochondrial membrane potential (Fig. 2B). These data suggest that a combination of IMT and venetoclax is cytotoxic and synergistically induces cell death in sensitive cells, presumably through disruption of mitochondrial function.

Combination of IMT and venetoclax significantly suppresses OXPHOS in AML cells

We next studied whether combination treatment with IMT and venetoclax affects OXPHOS in AML cells. OXPHOS was evaluated by determining oxygen consumption rate (OCR) (Fig. 3A), which was measured first as basal respiration and then oligomycin (OMY)-induced leak respiration (ATP-linked respiration) after the addition of oligomycin to inhibit ATP synthase. Stepwise carbonyl cyanide m-chlorophenyl hydrazone (CCCP) titrations were subsequently carried out to reach the electron transport chain respiratory capacity state. Finally, antimycin a (AMA) was added to inhibit complex I/III of the electron transport chain and to terminate respiration, allowing for the determination and correction of residual oxygen consumption, which is indicative of non-mitochondrial oxygen consumption (Fig. 3A). We noted that combination treatment significantly inhibits OXPHOS respiration in venetoclax-sensitive MOLM-13 and MV4-11 cells (Fig. 3B). In comparison, OXPHOS levels in venetoclax-resistant THP-1 cells were unaffected in the presence of single compound or combination treatments, confirming that the inhibition we observed is not a general effect of drug treatment (Fig. 3B). In addition, there was no conclusive OCR data obtained for KG-1 cell line.

Combination of IMT and venetoclax dramatically decreases OXPHOS (in a glycolysis-independent manner) in sensitive cell lines.

(A) Representative graph of an OXPHOS measurement using mitostress compounds: oligomycin (OMY), carbonyl cyanide m-chlorophenyl hydrazone (CCCP), and antimycin (AMA). (B) AML cell lines (MV4-11, MOLM-13, and THP-1) were treated with either IMT (500 nM), venetoclax (5 nM), or a combination of both for four days. DMSO was added to control wells. Oxygen consumption rate (OCR) was measured with an Oroboros O2K instrument at physiological glucose levels (5 mM) using mitostress compounds. (C) AML cell lines were cultured in growth medium supplemented with either 5 mM glucose or galactose, and cell viability tests (ATP measurements) were performed after 24 and 48 h. All results are from at least three independent experiments. Error bars show ± SEM. Statistical tests on all comparisons in (C) showed non-significant differences.

In normal hematopoietic cells, glycolysis is upregulated when OXPHOS is blocked (17). In contrast to normal hematopoietic cells, high energy turnover AML cells rely predominantly on OXPHOS due to poorly regulated metabolic adaptation. To verify the reliance on OXPHOS, AML cells were grown in growth medium supplemented with either glucose or galactose (Fig. 3C). Two ATP molecules are produced when cells use glucose for glycolysis, but if galactose is used instead, no ATP is generated (18). We observed no significant decrease in the level of ATP production (Fig. 3C) nor in cell counts (Fig. S3A) in the absence of glucose, indicating that glycolysis is not a major source of energy in AML cells. To test whether AML cells use glycolysis under treatment conditions, OCR was measured in the presence or absence of a high concentration of glucose (Fig. S3B). OCR measured for basal respiration was not significantly different with excess amounts of glucose, again indicating the inability of AML cells to switch to glycolysis when OXPHOS is disrupted.

Combination of IMT and venetoclax synergistically delays leukemia initiation in cell-derived leukemia mouse models

Following the confirmed synergy between IMT and venetoclax in AML cell lines, we proceeded to test the dual inhibition of mitochondrial transcription and BCL-2 in an AML cell-derived xenograft (CDX) transplantation mouse model. First, the pharmacokinetic profile of IMT was determined in plasma, liver, and bone marrow (BM) of mice, 1, 8, and 24 h post-administration (Fig. S4A). As described above (Fig. S1), the IMT mouse PK profile was superior to IMT1B. IMT was detected in all tissues at all time points when administered alone. To determine the concentration of IMT and venetoclax when administered in combination, plasma samples were then collected 8–11 h post-administration of single or combination formulations and the concentrations of IMT and/or venetoclax in samples were measured (Fig. S4B). Values obtained were all within the effective range, as combination treatment did not affect the individual compound concentrations (14). For xenograft experiments, NBSGW mice were intravenously transplanted with MV4-11 AML cells (Fig. 4A). Two weeks after transplantation, engraftment of AML cells was confirmed (evaluated by gating on hCD45+ population by flow cytometry in peripheral blood). Treatment then began with vehicle, 50 mg/kg IMT, 100 mg/kg venetoclax, or a combination of IMT (50 mg/kg) and venetoclax (100 mg/kg) (Fig. 4A), dosed as such based on previously published protocols (14, 18). The efficacy of treatments was determined by measuring the proportion of transplanted AML cells in peripheral blood and by survival analysis (Fig. 4B). In all treatment groups, the hCD45+ population dropped immediately after treatment, and vehicle- and venetoclax-treated mice then exhibited a rapid rebound. In contrast, in IMT-treated mice the increase in leukemic cell proportion was delayed until eight weeks post-transplantation, and in the combination-treated group for to up to 20 weeks (Fig. 4B). These observations correlate with results from survival analysis (Fig. 4C). IMT treatment of MV4-11 transplanted mice significantly delayed the development of leukemia, inducing a 37% increase in median overall survival (57.5 → 79 days). Development of leukemia was further delayed by combination treatment to 131% (57.5 → 133 days), and the group contained two mice that did not show any disease symptoms after 200 days (post-transplantation) when they had to be euthanized according to ethical permissions. Moreover, leukemic cells were not detectable in the BM of surviving mice at the time of termination. All other mice showed symptoms of leukemia and of these, all but one combination-treated mouse had an increased percentage of leukemic cells in BM (Fig. 4D). These results confirm that treatment with a combination of IMT and venetoclax significantly prolongs leukemia progression in mice. Moreover, the general condition of mice such as behavior, fur, mobility and weight, assessed before (week 2) and after treatment (week 5), were not altered in the IMT-, venetoclax-, or combination-treatment groups when compared to the vehicle-treated group (weights shown in Fig. S5B), suggesting that the treatments were well tolerated. As with single IMT therapy (14), treatment with combination therapy did not display any evidence of organ toxicity or drug-associated death in the mice (Fig. S5C, D).

Combination of IMT and venetoclax significantly prolongs the survival of mice with leukemia.

(A) Outline of CDX efficacy experiment: NBSGW mice were transplanted with MV4-11 human leukemia cells, then treated with either vehicle, IMT, venetoclax, or a combination of IMT and venetoclax (n=10 per treatment group). Peripheral blood was taken before treatment (week 2) and after treatment at regular intervals until disease progression was observed. (B) Analysis of proportion of human CD45+ cells in mice peripheral blood in all treatment groups at indicated time points, determined by flow cytometry. (C) Kaplan-Meier analysis of survival. Statistical significance was calculated using log-rank (Mantle-Cox) test. ****p < 0.0001, Error bars represent ± SEM. (D) Analysis of proportion of human CD45+ cells in the bone marrow of all treatment groups (four mice per group compared to two surviving mice) at time of euthanization.

IMT and venetoclax synergistically decrease the tumor burden and increase the life span in an AML-PDX mouse model resistant to venetoclax

The CDX mouse model demonstrated IMT could sensitize AML cell lines to BCL-2 inhibitors in vivo. We next analyzed the effects of combination treatment on a more refined in vivo model by generating xenografts using AML patient-derived cells (AML PDX). We first tested an AML-PDX mouse model with resistance to venetoclax (AML sample harboring TP53 mutation). Two weeks post-transplantation with AML PDX cells, NSGS mice were orally treated with IMT, venetoclax, or a combination of the two for 21 days (Fig. 5A). The efficacy of the treatments was determined based on the proportion of human AML cells (determined by hCD45+ population). Six out of 10 mice in each group were euthanized immediately after the last compound administration for end-of-treatment analysis. We found that the combination of IMT and venetoclax resulted in significant inhibition of tumor burden in blood, BM, and spleen of treated mice (Fig. 5B). By examining histology sections of BM and spleen, we found that tissues from combination-treated animals contained less engraftment of leukemic blasts than single- and vehicle-treated mice (Fig. 5C). Accordingly, the frequencies of hCD45+ cells in the BM and spleens of the combination-treated group were largely reduced in comparison to other groups (Fig. 5D), and engrafted tumor cells in combination-treated PDX mice showed a lower proliferating percentage in comparison to single or vehicle treated groups (shown by KI67+/hCD45+ immunohistochemistry staining; Fig. 5E). This demonstrates a synergy between IMT and venetoclax in reducing tumor burden.

Synergistic effect of (IMT and venetoclax) combination treatment on tumor burden and survival of AML-PDX transplanted mice.

(A) Experimental outline: NSGS mice were transplanted with primary AML splenocytes from PDX mice and were then treated with either vehicle, IMT, venetoclax, or a combination of IMT and venetoclax (n=10 per treatment group). After the final dosing, three or four mice were euthanized (end-of-treatment group) and the rest (n=6) were maintained for survival analysis. (B) Percentage of viable human CD45-positive (hCD45+) cells in peripheral blood, BM, and spleen of treated mice determined by flow cytometry. (C) Immunohistochemistry analysis of mouse BM and spleen sections stained with anti-human CD45 (brown) and counterstained with DAB (blue). (D) Quantitative analysis of hCD45+ cells on immunohistochemistry sections of BM and spleen sections of treated mice. (E) Proliferating tumor cells (CD45) are shown by quantification of Ki67 positive cells in immunohistochemistry-stained BM sections at end-of-treatment (results from at least three technical replicates and two biological replicates in each treatment group).

To further analyze the cellular effects of the IMT-venetoclax combination treatment, we analyzed transcriptomes of purified viable AML blasts isolated from PDX models at end of treatment (Fig. 6A). We confirmed that combination treatment had a more pronounced effect on gene transcription as compared to single treatments. In agreement with its more pronounced effect on tumor burden, combination treatment suppressed expression of a number of genes involved in leukemogenesis, including MX1, KCTD12, ITGA and RGL1 (Table S2) (19).

Synergistic effect of (IMT and venetoclax) combination treatment on gene transcription and survival of AML-PDX transplanted mice.

The experimental outline was shown in figure 5A, but n=10 per treatment group. After the final dosing, three or four mice were euthanized (end-of-treatment group) and the rest (n=6) were maintained for survival analysis. (A) Gene expression profiling by RNA-seq of purified AML blasts isolated from PDX models treated with venetoclax, IMT, or a combination thereof. Indicated expression changes are relative to a vehicle treated group. Significantly enriched genes are indicated in red (q < 0.05). (B) Peripheral blood engraftment analysis of survival group. (C) Survival analysis of treated mice shown by Kaplan-Meier analysis (n=6/treatment group). *P < 0.05, **P < 0.01, and ***p < 0.001.

We also analyzed survival by monitoring the mice for disease progression and morbidity. Peripheral blood engraftments were analyzed at regular intervals until the appearance of disease symptoms (Fig. 6B and Fig. S6A). The leukemia burden immediately prior to euthanization demonstrated that the cytoreductive effect of the combination treatment persists (Fig. 6B). However, blood count (determined by RBCs and HGB levels) did not vary significantly between experimental groups, nor did the magnitude of splenomegaly or liver weight vary at the time of euthanization (Fig. S6B–D). Interestingly, mice treated with a combination of IMT and venetoclax succumbed to death significantly later than other treatment groups, with a median survival of 112 days vs 96, 99, and 102 days for vehicle-, IMT-, and venetoclax-treated groups, respectively (Fig. 6C). Combination treatment thus increases the life span of venetoclax-resistant AML mouse models, demonstrating that IMT can sensitize AML cells towards other therapies.

IMT does not affect tumor burden or life span in an AML-PDX mouse model sensitive to venetoclax

For comparison, we also monitored effects of combination treatment on a PDX model based on AML cells sensitive to venetoclax treatment (AML cells harboring FLT3-ITD, NMP1 and DNMT3A). As expected, we observed that venetoclax treatment significantly reduced the tumor burden in the blood and BM of mice. Combination treatment with IMT did not add to this effect (Fig. 7A). By examining histology sections, we found that both BM and spleen tissues from combination- and venetoclax-treated animals contained reduced proportion of leukemic blasts compared to IMT- and vehicle-treated mice (Fig. 7B). White blood counts (WBCs), red blood counts (RBCs), and hemoglobin (HGB) levels were not significantly different across experimental groups (Fig. S7A). The total body and liver weight of PDX mice were very similar in all groups (Fig. S7B, D), but both venetoclax and combination treatments reduced splenomegaly (Fig. S7C).

Leukemia progression induced by primary AML cells is inhibited by venetoclax treatment.

As in Fig. 5A, efficacy experiment outline: NSGS mice were transplanted with primary AML splenocytes from PDX mice and were then treated with either vehicle, IMT, venetoclax, or a combination of IMT and venetoclax (n=12 per treatment group). 8–12 h after the last dosing, half of the mice (n=6/treatment group) were euthanized (end-of-treatment group) and the other half were maintained for survival analysis. (A) Percentage of viable human CD45-positive (hCD45+) cells in peripheral blood and bone marrow of treated mice (n=6 for each treatment group), determined by flow cytometry. (B) Immunohistochemistry analysis of hCD45+ cells on bone marrow and spleen tissue sections of treated mice (images are representative from at least two mice per group). Scale bars represent 100 μM. (C) Peripheral blood engraftment analysis of survival group. (D) Mice treated with single or a combination of compounds showed prolonged survival in comparison to the vehicle-treated group. Kaplan-Meier analysis shows the relative survival curve. *P < 0.05, **P < 0.01, and ***p < 0.001.

The remaining half of mice were used for survival analysis, which involved regular blood count and chimerism analysis until the appearance of disease symptoms (Fig. 7C). Both venetoclax and IMT caused a modest increase in life span, but combination treatment did not offer an additional effect (Fig. 7D).

Discussion

In recent years, the genetic complexity and heterogeneity of leukemia cell populations has become apparent (20, 21) and developing effective strategies to target human leukemia initiating cells has thus proven to be a challenging task. Currently, there is considerable interest in therapies, such as venetoclax, that can restore normal apoptosis of cancer cells by directly interacting with the BCL-2 family of proteins. These are effective for some leukemias, but many patients eventually become refractory to treatment and ultimately succumb to their disease (22, 23). It has become increasingly important to determine which biological properties are relatively consistent in leukemic cell populations irrespective of genetic makeup (tumor population). Recent observations indicate that several cancer cell types, including AML, are reliant on OXPHOS for metabolism (24), making inhibition of mitochondrial function a promising therapeutic strategy for AML (12, 25–28). Therapy-resistant leukemic cells are more reliant on oxidative respiration than glycolysis for energy generation, suggesting that the OXPHOS dependency and lack of normal glycolytic activity in AML cells may be characteristic of cancers (29). Understanding resistance mechanisms will allow for the design of strategies to counteract drug resistance or impede resistance from developing. In AMLs with aberrant expression of BCL-2, venetoclax can reduce OXPHOS (9) and trigger cell death. However, venetoclax resistance can occur in patients treated with the BCL-2 inhibitor and in some subtypes of AML patients with poor prognoses (30).

Previous work has indicated that AML cells display increased mitochondrial biogenesis and inhibition of mitochondrial translation can be used as a strategy for human AML (18, 27). In our work, we target another process involved in mitochondrial gene expression, i.e., transcription. IMT treatment targets POLRMT, which has two main functions in mitochondria: it transcribes mtDNA encoded genes and provides the primers required to initiate mtDNA replication. Inhibition of POLRMT can thus deliver a double punch, both reducing mtDNA copy number and inhibit OXPHOS biogenesis. As demonstrated here inhibition of POLRMT can overcome venetoclax therapy resistance in AML. These observations are in line with recent reports of the importance of mitochondrial function in leukemia cell survival (31, 32). By developing a clinically relevant and well-tolerated approach to deliver IMTs in immunodeficient mice transplanted with AML cells, we were able to reduce the tumor burden in all (cell- and patient-derived) xenografted mice treated with a combination of IMT and venetoclax. This demonstrates that reducing POLRMT activity results in successful targeting of leukemic cells. Our in vitro investigations confirmed that the combination of IMT and venetoclax is necessary to significantly decrease ATP levels and consequently induce apoptosis. In addition, efficacy studies of AML PDX models showed that single compounds are not sufficient to eradicate AML cells, but that the combination regimen reduced AML tumor burden and prolonged survival in all treated mice. These data indicate the advantage of both downregulating POLRMT activity in combination with increasing apoptosis and highlights the importance of rational combination therapies (33, 34).

In summary, we have introduced a new therapeutic strategy to target therapy-resistant leukemia with OXPHOS-dependent properties. Despite several limitations with the design of pre-clinical cancer models and the intensity and duration of treatments, we demonstrate here that the combined inhibition of BCL-2 and mitochondrial transcription can target therapy-resistant AMLs. This combination treatment succeeds as IMT primes cell vulnerability to apoptosis induction. Consequently, our findings may be of value for the future development of better clinical strategies for the treatment of AML.

Supporting information

eLife Assessment

This solid study assesses a novel mitochondrial inhibitor in combination with the BCL-2 inhibitor venetoclax, with the aim to increase its activity in acute myeloid leukemia. It provides valuable findings of combinatorial efficacy using preclinical models, confirming the overall importance of targeting oxidative phosphorylation to overcome venetoclax resistance in acute myeloid leukemia, and could be strengthened through mechanistic studies demonstrating drug specificity, pharmacodynamic efficacy studies in vivo to test clinical utility and extended statistical analyses of the results. The study is of interest to hematologists because it addresses a key biomedical issue in acute myeloid leukemia (venetoclax resistance) and provides data regarding the safety and activity of a novel inhibitor of the mitochondrial polymerase addressed in combination with venetoclax.

Reviewer #1 (Public Review):

This study exploits novel agent (IMT) that inhibits mitochondrial activity in combination with venetoclax. While the concept is not novel, the agent is novel (inhibitor of the mitochondrial RNA polymerase, described in Nature in other tumor models), and quest for safe mitochondrial inhibitors is highly warranted. The strength is in vivo activity data shown in CLDX and in one of the two AML PDX models tested, and the apparent safety of the combination. However, the impact on survival is impressive in CLDX but not in PDX, and unclear why Ven-sensitive PDX is resistant to combination (opposite what cell line data show). The paper is lacking mechanistic data beyond Seahorse and standard apoptosis assays, and even transcriptome analysis from PDX cells is poorly analyzed. There is no real evidence that this agent overcome Ven resistance, which could be done for example in primary AML cells. Finally, no on-target pharmacodynamic endpoints are measured in vivo to support the activity of the compound on mitochondrial activity at the doses used (which are safe). These multiple weaknesses significantly reduce my enthusiasm for this manuscript.

The cell line data show additive/synergistic effects of IMT and Ven on cell viability in p53-WT cells. However, no mechanisms of synergy beyond OCR are shown, which is a missed opportunity.

No data are shown in primary AML cells in vitro. This could address venetoclax-resistant AML cells with distinct genomic profiles.

The in vivo CLDX model (MV4;11) data is quite impressive, showing reduction of tumor burden and meaningful extension of survival in combination cohort. It is unclear why venetoclax used at highest dose normally sued in vivo (100mg/kg) did not show any impact on survival in this Ven-sensitive model. It is disappointing that no biomarkers of mitochondrial activity (for example, simple pAMPK, or levels of mitochondrial subunits) are shown to support on-target pharmacodynamic activity. However, efficacy in human PDX is less impressive, for example in Fig 6C the combination has extended survival from 96 to 112 days, possibly due to early stopping of treatment (around day 30); and no extension of survival is seen in another PDX in Fig 7. Still, this is indicative of combinatorial activity in TP53-mutant PDX. There is however discrepancy with in vitro studies that show no impact of combination in TP53 mutant cells and synergy in TP53-wt cells, and the opposite findings in vivo, which is not explained. Overall, the activity of the combination is modest. The safety is encouraging, but again, no pharmacodynamic measurements are shown to support that IMT at least partially inhibited mitochondrial activity in AML cells.

In Discussion the statement that inhibition of POLRMT can overcome venetoclax resistance is not supported by the data, as no additive effects are seen in vitro in TP53 mutant cells, and no other resistant models (such as primary AML cells) are tested. In vivo as stated above there is some activity in TP53 mutant PDX but this alone cannot be sued to justify this strong statement. Also, the sentence that "...we were able to reduce the tumor burden in all (cell- and patient-derived) xenografted mice treated with a combination of IMT and venetoclax" is not supported by data in Fig 7.

Reviewer #2 (Public Review):

Summary:

The manuscript by Arabanian and colleagues presents studies showing how inhibition of mitochondrial transcription and replication with a novel inhibitor of the mitochondrial polymerase, IMT, can promote AML cell death in combination with the Bcl2 inhibitor venetoclax. They further show that this combinatorial efficacy is evident in vivo in both the AML cell line MV411 and in a PDX model. Given the multiple studies showing the importance of Oxphos in maintaining AML cell survival, the current studies provide an additional strategy to inhibit Oxphos and thus improve the therapeutic management of AML.

Strengths:

A novel aspect of this work is that IMT is a new class of mitochondrial inhibitor that acts by inhibiting the mitochondrial polymerase. In addition, the demonstration of therapeutic efficacy both in vitro and in vivo (including with PDX), together with some data showing minimal toxicity, adds to the impact of this work. Their overall conclusion that IMT increases the potency of Vex in treating AMLs is supported.

Weaknesses:

There are several deficiencies that should be addressed to substantiate the rigor and impact of this study. Of most importance, they need to show that IMT actually inhibits the mitochondrial polymerase in AML cells, and there are additional concerns with their models that if addressed would improve the ability of IMT to be developed clinically.

Author response:

The following is the authors’ response to the original reviews

Reviewer #1:

(1) The data are generated using ATP read-out (CTG assay). For any inhibitor of mitochondrial function, ATP assays are highly sensitive reflecting metabolic stress, yet these do not necessarily translate into cell growth inhibition using standard Trypan blue assays and tend to overestimate the effects. Please show orthogonal more robust assays of cell growth or proliferation.

We acknowledge the sensitivity of the ATP read-out assay in reflecting metabolic stress. While additional cell growth assays such as Trypan blue exclusion could provide further insights, we believe that the current ATP assay data robustly demonstrate the effect of the IMT and venetoclax combination on cellular metabolism, which is a critical aspect of our study. The scope of our current work focused on metabolic inhibition, and we suggest that future studies could further explore cell proliferation assays to complement these findings.

(2) It is concluded that AML cells do not utilize glucose for ATP production. Please provide formal measurements of glycolysis/lactate upon combinatorial treatment.

We appreciate the reviewer’s suggestion to include glycolysis and lactate measurements, which could indeed add further granularity to our metabolic analysis. However, the primary focus of our study is on mitochondrial function and oxidative phosphorylation (OXPHOS) in AML cells treated with IMT and venetoclax. We believe the data presented in Figure 3 provide strong support for the conclusion that glycolysis is not a major energy source in these cells.

Specifically, in Figure 3C, we demonstrate that AML cells maintain ATP levels and viability when cultured in galactose, a condition that restricts ATP production through glycolysis and forces cells to rely on OXPHOS. This result strongly suggests that these AML cells are not dependent on glycolysis for ATP production. Furthermore, in Supplementary Figure S3B, we show that oxygen consumption rate (OCR) measurements remain stable in the presence of excess glucose, further supporting our conclusion that the cells do not switch to glycolysis when OXPHOS is inhibited.

These findings collectively indicate a primary reliance on OXPHOS for energy generation in AML cells, consistent with our study’s objectives to explore mitochondrial dependency and the therapeutic potential of targeting mitochondrial transcription in AML. Future studies could certainly expand on these insights by incorporating a more detailed analysis of glycolytic flux and lactate production under combinatorial treatment, but we believe the current data are sufficient to support our main conclusions.

(3) The transcriptome data are shown without any analysis of pathways. The conclusion from this data beyond the higher number of genes impacted in the combination arm is unclear. Please provide analysis for example GO pathways and interpret in the context of the drugs' mechanism of action.

In response to the reviewer’s question, we have added gene ontology (GO) pathway analysis to clarify the transcriptomic impact of our combination treatment with IMT and venetoclax. Functional annotation identified significant enrichment in pathways relevant to innate immune response, mitochondrial function, and cellular signaling processes. Specifically, pathways associated with immune defense, mitochondrial signaling, and intracellular signaling were notably affected. These findings suggest that the combination treatment not only disrupts cellular energy metabolism but also potentially primes immune signaling mechanisms. This aligns with the proposed mechanism, where IMT targets mitochondrial transcription and venetoclax induces apoptosis, together enhancing sensitivity in AML cells. The enriched pathways, therefore, support the mechanism of action of both drugs, showing how the combined inhibition of BCL-2 and mitochondrial transcription creates a compounded cellular disruption that enhances the therapeutic effect.

(4) Please demonstrate (could be in supplement) matrix of combination to support the statement that the combination is synergistic using Bliss index. The actual Bliss values are missing.

For the revision, we have now included a matrix of combination treatment effects with the corresponding Bliss synergy index values to substantiate our claim of synergy between IMT and venetoclax. This analysis, provided in the supplement, demonstrates that the observed effects exceed the expected additive impact of each drug alone, as calculated by the Bliss independence model. Specifically, the Bliss values confirm a synergistic interaction in venetoclax-sensitive AML cell lines, highlighting that the combined treatment significantly enhances inhibition of cell viability and apoptosis induction compared to single treatments. This data supports our interpretation of synergy and strengthens the mechanistic conclusions drawn from our findings on the combination therapy’s efficacy.

(5) Please show KG1 data (OCR), here or in Supplement.

In response to the reviewer’s request to include OCR data for the KG-1 cell line, we would like to clarify that OCR measurements were attempted; however, they did not yield conclusive results. This is noted in the revised manuscript (Results section), where we explain that the KG-1 cell line did not provide usable OCR data, likely due to limitations in detecting reliable mitochondrial respiration in this particular line under our experimental conditions. Therefore, we were unable to include KG-1 OCR data in the main figures or the supplement.

Reviewer #2:

(1) It's important that the authors show that the drug's effects in AML are due to on-target inhibition. It's critical that they show that IMT actually inhibits the mito polymerase in the AML cells in the dose range employed.

We appreciate the importance of demonstrating on-target inhibition of mitochondrial RNA polymerase by IMT1, especially in light of the detailed characterization of IMT1b, a closely related compound, as presented in Bonekamp et al., Nature 2020. The work by Bonekamp et al. established the specificity and efficacy of IMT1b in targeting mitochondrial RNA polymerase across various tumor models. Building on these findings, we designed our study to primarily evaluate the combinatorial efficacy of IMT1 with venetoclax in AML models, assuming a similar mechanism of action as described for IMT1b. While direct confirmation of on-target inhibition in AML cells by IMT1 would undoubtedly provide additional mechanistic insight, we focused on translational aspects in this study. We believe that the foundational work provided by Bonekamp et al. supports the assumption of on-target effects by IMT1, and we suggest that future studies could explicitly verify this in the context of AML.

(2) For Fig 1, the stated synergism between Venetoclax (Vex) and IMT in p53 mutant THP1 cells is really not evident, despite what the statistical analysis says. In some ways, the more interesting conclusion is that inhibiting mitochondrial transcription does NOT potentiate the efficacy of Bcl2 inhibition in TP53 mutant AML.

We appreciate the reviewer’s observation regarding the lack of evident synergy between IMT and venetoclax in TP53 mutant THP-1 cells. In line with this comment, we have now expanded the discussion to emphasize that, while statistical analysis suggested a potential interaction, the biological response in TP53 mutant cells was minimal. This contrasts with the strong synergy observed in TP53 wild-type cell lines, such as MV4-11 and MOLM-13. We have now highlighted that TP53 mutation status may limit the effectiveness of mitochondrial transcription inhibition in potentiating BCL-2 inhibition. This addition underscores the importance of mutation profiles, such as TP53 status, in predicting response to combination therapies in AML and is now clearly addressed in the revised discussion.

(3) They combine IMT with Vex, but Vex plus azacytidine or decitabine is the approved therapy for AML. Any clinical trial would likely start with this backbone (like Vex+Aza). They should test combinations of IMT with Vex/Aza or Vex/Dec.

While we recognize the importance of testing IMT in combination with clinically approved therapies like Vex+Aza, our current study was designed to explore the potential of IMT in combination with venetoclax alone. Expanding to other combinations would be an excellent direction for future research but is beyond the scope of our current investigation.

(4) It's interesting that AML cell lines do not show any reliance on ATP generation from glycolysis, but would this still be the case when OxPhos is inhibited with IMT? Such a simple experiment would be much more interesting and could help them better understand the mechanism of IMT efficacy.

We thank the reviewer for highlighting this point regarding the reliance of AML cell lines on glycolysis under OxPhos inhibition. In our study, we observed that AML cells predominantly rely on OxPhos, and we did test for ATP production in conditions that favored glycolysis by growing AML cells with galactose instead of glucose in the medium. As described in the manuscript, we did not observe significant ATP production or cell viability from glycolysis, even under these conditions. This finding suggests that AML cells have a low capacity to adapt to glycolytic ATP generation when OxPhos is disrupted by IMT, reinforcing the view that they are highly dependent on mitochondrial function for energy production. We agree that this adaptation—or lack thereof—is an intriguing aspect of IMT efficacy in targeting energy metabolism in AML cells, and we have clarified this point in the discussion.

(5) OxPhos measurements need statistical analyses.

We appreciate the reviewer’s suggestion to include statistical analyses for the OXPHOS measurements. We would like to clarify that statistical analyses were included in the initial submission. These are detailed in Figure 3 and its legend, as well as in the Statistical Analysis section, where we specify methods such as the calculation of standard error across replicates. This approach was implemented to ensure the rigor of our OCR data and its conclusions on OXPHOS inhibition in AML cells.

(6) Given that the combo-treated mice do not exhibit much leukemia in the blood through ~180 days, and yet start dying after 100 days, the authors should comment on this, given that the bone marrow has been shown to be a refuge that protects leukemia cells from various therapies.

We thank the reviewer for highlighting the observed discrepancy between peripheral blood leukemia levels and survival in combo-treated mice. While leukemic cells were minimally detected in the blood up to approximately 180 days, treated mice began to show signs of disease progression and reduced survival around 100 days. This may suggest that residual leukemic cells persist within the bone marrow, which has been established as a sanctuary site for leukemic cells, providing protection from various therapies. The bone marrow environment likely supports a survival niche, enabling these residual cells to evade treatment effects and potentially initiate disease relapse. We have added this interpretation to the discussion to acknowledge the possibility of bone marrow as a protective refuge, which may limit the full eradication of leukemia in these models despite apparent peripheral blood clearance.

(7) For Fig 5C, the authors should statistically compare the Combo with Vex alone.

We have now included statistical comparisons between the combination treatment and venetoclax alone in Fig 5C to provide a clearer interpretation of the data.

(8) The analyses of gene expression using RNAseq of harvested leukemia cells from the PDX model (Table S2), some more discussion of these results would be helpful, particularly given that neither drug is directly targeting nuclear gene expression.

We thank the reviewer for their suggestion to discuss the RNAseq findings in more detail. In the revised manuscript, we have expanded on the functional annotation of the gene expression changes observed in leukemia cells from the PDX model following combination treatment (Table S2). The enriched pathways include innate immune involvement, mitochondrial function and immune signaling, and intracellular signaling. This suggests that while neither IMT nor venetoclax directly targets nuclear gene expression, the combined treatment induces secondary effects that alter these pathways, potentially contributing to the treatment’s efficacy in AML. This expanded discussion provides greater insight into how the drug combination impacts gene expression and cellular pathways.

(9) We need more information on the PDX models, in terms of the classification (M1 to M6) of the patient AMLs and genetics (specific mutations, not just the genes mutated, and chromosomal alterations).

Additional details regarding the classification and genetic background of the PDX models have been included in the manuscript to better contextualize our findings.

(10) The authors should discuss whether or not IMT represents an improvement over other therapies intended to target Oxphos in AML (clearly, the low toxicity of IMT is a plus, at least in mice).

We appreciate the reviewer’s suggestion to discuss IMT in comparison with other OXPHOS-targeting therapies for AML. In the revised discussion, we highlight IMT’s unique properties, particularly its low toxicity profile, which may offer advantages over other OXPHOS inhibitors. This low toxicity, demonstrated in preclinical studies, suggests that IMT might improve patient tolerability compared to existing therapies that target mitochondrial function.

(11) The authors examined toxicity by weighing the mice and performing CBCs. Measurements of liver and kidney toxicity will be necessary for further clinical development.

We thank the reviewer for the suggestion to further investigate liver and kidney toxicity. In our study, we assessed toxicity through regular weight monitoring and complete blood counts (CBCs) to evaluate overall health status. While additional liver and kidney toxicity measurements will indeed be important in future studies, resource limitations currently prevent us from performing these additional analyses in this model. We agree that these assessments will be essential as we progress towards clinical development, and we plan to address them in upcoming preclinical studies.

Version 2

Abstract

Abstract

Overcoming drug-resistance and the subsequent relapse that often occurs with monotherapy is crucial in the treatment of acute myeloid leukemia. We here demonstrate that therapy-resistant leukemia initiating cells can be targeted using a novel inhibitor of mitochondrial transcription (IMT). The compound inhibits mitochondrial RNA polymerase activity and sensitizes the resistant population to the induction of apoptosis. In vitro studies on acute myeloid leukemia cells demonstrate that IMT prevents cell proliferation, and together with a selective BCL-2 inhibitor, venetoclax, induces apoptosis and suppress oxidative phosphorylation (OXPHOS) synergistically. AML mouse models treated with IMT in combination with venetoclax show prolonged survival in venetoclax-resistant models. Our findings suggest that certain therapy-resistant leukemia cell populations display a unique dependency on mitochondrial transcription and can be targeted with IMT.

Introduction

Acute myeloid leukemia (AML) comprises a devastating and heterogeneous group of diseases, for which patient survival rates are dismal despite several new treatment options. Current cares are still mainly based on unselective chemotherapy but for most patients also with addition of more directed treatments after patient stratification based on genetic aberrations found at diagnosis. However, these strategies lead to cures in only a fraction of AML patients due to therapy resistance developed by the leukemic cells leading to relapse. Curative treatment usually needs allogenic stem cell transplantation, which is not applicable in many cases, highlighting the critical need to develop different approaches. The anti-apoptotic molecule BCL-2 has been a therapeutic target in lymphoid and myeloid leukemia, as in many cancers overexpression of BCL-2 results in dysregulation of apoptosis (1). ABT-199 (venetoclax), is a specific BH3-mimetic that causes apoptosis by preventing BCL-2 from binding pro-apoptotic BAX and BAK1 proteins (2). However, complete responses occur in just ∼20% of AML patients treated with Venetoclax (3). In addition, many patients relapse, with regrowth initiated by therapy-resistant leukemic clones, suggesting that single agent administration is unlikely to yield lasting responses in most cases. Venetoclax is therefore used in combination with conventional therapies. Together with hypomethylating agents, Venetoclax is now standard of care for patients deemed unfit for intensive induction therapy (4).

Resistance to venetoclax may result from overexpression of the anti-apoptotic proteins MCL1 or BCL2L1 (5), or through changes in levels of other anti-apoptotic molecules (6). Aberrant pro-survival regulation can also be achieved by inactivation of TP53 function, which controls many aspects of apoptosis (6, 7). MCL1-dependent resistance to venetoclax can be overcome when the compound is combined with the MDM2 inhibitor RG7388, which causes p53 activation. A combination of BCL-2 inhibition and TP53 activation may thus provide an interesting therapeutic approach for treatment of AML (8).

Other combination approaches that promote apoptosis may also be of interest. The majority of ATP required for leukemia stem cell (LSC) function is generated by oxidative phosphorylation (OXPHOS) (9, 10). Interestingly, venetoclax-resistant leukemic cells display increased levels of OXPHOS and are protected from mitochondrial stress (7). These data suggest approaches to specifically target OXPHOS in leukemia may be an attractive option. In support of this notion, inhibition of mitochondrial protein translation by antibiotics that reduce OXPHOS capacity results in a selective cytotoxic effect on leukemic cell when used in combination with tyrosine kinase inhibitors (11). Furthermore, LSCs from de novo AML patients display a unique dependency on amino acid metabolism to fuel OXPHOS. When venetoclax is combined with the hypomethylating agent azacytidine, the two compounds act in concert to inhibit amino acid uptake, leading to a drop in OXPHOS and thus cell death (12). More recently, AML therapy using a combination of cytarabine (AraC) and venetoclax has highlighted the central role of mitochondrial adaptation during treatment (10, 13), indicating the need for a mitochondrial directed combination therapy for AML. Such a therapy may help to overcome existing treatment resistance.

Recently we developed novel, first-in-class inhibitors of mitochondrial transcription (IMTs) (14). IMT treatment specifically targets mitochondrial RNA polymerase (POLRMT) and reduce transcription of mitochondrial DNA, which codes for thirteen key components of the OXPHOS machinery, including subunits of respiratory complexes I, III, IV, and the ATP synthase. Consequently, IMT treatment impairs OXPHOS system biogenesis in a dose-dependent manner and decreases cell viability in various tumor cells. Surprisingly, several weeks of per oral IMT treatment is well tolerated in mice and does not cause toxicity in normal tissues, despite inducing a strong anti-tumor response in human cancer xenografts (14).

Here we examine AML cell lines for their sensitivity for treatment with IMT. We find that a combination of IMT and venetoclax displays a synergistic on cell growth and induction of apoptosis. We demonstrated the efficacy of this approach in leukemia mouse models generated by human AML cell lines as well as primary samples from AML patients.

Results

Synergistic effect of IMT in combination with venetoclax on AML cell lines

We were curious whether our inhibitors of mitochondrial transcription (IMTs) (14) could be effective against AML cells. The coumarin class of IMTs was optimized by medicinal chemistry to generate lead IMT compounds with improved metabolic stability in mice, such as LDC204857, herein simply referred to as IMT (Fig 1A, Fig. S1 and Table S1). IMT is a close analog of previously described IMT1B (Fig. S1) with a highly similar biochemical and cellular inhibition profile (14). It shows an extended half-life in vivo due to its increased metabolic stability (Table S1). Thus, by using IMT we could consistently reduce the mouse dose to 50 mg/kg per oral once daily, which is very well tolerated and does not exert any adverse effects in mice.

IMT+venetoclax combination treatment synergistically inhibits cell viability/growth of AML cell lines in vitro.

(A) Chemical structure of IMT (LDC204857) (B) AML cell lines were treated with single agents IMT or venetoclax at a range of concentrations (0.1–1000 nM and 0.01–100 nM, respectively), and assessed by cell viability tests after three and six days. (C) Synergism analysis of AML cell lines treated with a combination of IMT (0.05, 0.1, 0.2 μM for MV4-11 and MOLM-13, and 0.1, 0.5, and 1 μM for THP-1 and KG-1) and venetoclax (Vex; 1, 2, 4 nM for MV4-11 and MOLM-13, and 1, 5, 10 nM for THP-1 and KG-1) for six days. ‘S’ represents synergy of the combination treatment, evaluated using the Bliss independent model. Proliferation percentages are normalized to DMSO treated. All results are from three independent experiments.

Using four AML cell lines with different expression of TP53 (MV4-11 and MOLM-13 expressing wild type TP53; THP-1 and KG-1 expressing mutated TP53), we monitored the effects of increasing concentrations of IMT on cellular growth using a luminescence-based cell viability assay. Only marginal effects were observed after a 72-h period of exposure to IMT, but all cell lines displayed dose-dependent antileukemic activity after 144 h of treatment (Fig. 1B), in line with our previous observations (14). We also monitored venetoclax sensitivity by assessing cell viability, and noted that, in line with published work (15), a dose-dependent reduction in cell growth occurs after three days and becomes more pronounced after 6 days (Fig. 1B).

To then investigate whether IMTs show a synergistic effect when combined with venetoclax, we conducted a combination study in AML cell lines (Fig. 1C). After six days of treatment, we found a clear synergistic effect on cell proliferation in MV4-11 and MOLM-13 cell lines, which express wild type TP53, as determined by combination analysis (Bliss-independence model (16)). Treatment of THP-1 and KG-1 cell lines, which harbor mutated TP53, does not show a synergistic effect even when using higher concentrations of both IMT and venetoclax (Fig. 1C, additional treatment data are shown in Fig. S2, Bliss value calculations for both time points are summarized Fig. 1 - Source Data 1.

Combination of IMT and venetoclax synergistically induces apoptosis and loss of mitochondrial membrane potential in MV4-11 and MOLM-13 cells

To investigate the factors involved in the inhibition of cell growth with IMT and venetoclax, AML cell lines were treated with single agents (IMT or venetoclax) or a combination of the two (Fig. 2A), and the stages of apoptosis were monitored using Annexin V and PI staining. No detectable increase in apoptosis was induced by IMT exposure alone in any AML cell line. Venetoclax treatment resulted in both early and late apoptosis induction in MV4-11 and MOLM-13 cell lines, and employing a combination of IMT and venetoclax significantly enhanced apoptosis induction in these wild-type cell lines. On the other hand, combination treatment did not induce apoptosis in THP-1 and KG-1 mutated cell lines (Fig. 2A).

Combination treatment significantly enhances apoptosis induction in AML cell lines.

AML cell lines were treated with IMT (500 nM) and venetoclax (5 nM) for 72–96 h. (A) Apoptosis was determined by Annexin V/PI staining of the cells and analyzed by flow cytometry. (B) Loss of mitochondrial membrane potential was calculated as the percentage of depolarized (apoptotic) cells out of the total cell population. Error bars represent standard error of the mean where * represents P < 0.05, **P < 0.01, and ***p < 0.001.

One distinctive feature of apoptosis is the disruption of mitochondrial function, which can be measured as loss of mitochondrial membrane potential due to membrane depolarization. As a combination of IMT and venetoclax could synergistically induce apoptosis, we were curious whether combination drug treatment thus affects mitochondrial membrane potential. We observed that venetoclax-sensitive cell lines (MV4-11 and MOLM-13) showed a dramatic loss of mitochondrial membrane potential when treated with both venetoclax and IMT (Fig. 2B). On the other hand, mitochondrial membrane potential in THP-1 cells was not affected by single nor by combination treatment. Interestingly, although KG-1 was venetoclax insensitive, the combination treatment significantly enhanced loss of mitochondrial membrane potential (Fig. 2B). These data suggest that a combination of IMT and venetoclax is cytotoxic and synergistically induces cell death in sensitive cells, presumably through disruption of mitochondrial function.

Combination of IMT and venetoclax significantly suppresses OXPHOS in AML cells

We next studied whether combination treatment with IMT and venetoclax affects OXPHOS in AML cells. OXPHOS was evaluated by determining oxygen consumption rate (OCR) (Fig. 3A), which was measured first as basal respiration and then oligomycin (OMY)-induced leak respiration (ATP-linked respiration) after the addition of oligomycin to inhibit ATP synthase. Stepwise carbonyl cyanide m-chlorophenyl hydrazone (CCCP) titrations were subsequently carried out to reach the electron transport chain respiratory capacity state. Finally, antimycin a (AMA) was added to inhibit complex I/III of the electron transport chain and to terminate respiration, allowing for the determination and correction of residual oxygen consumption, which is indicative of non-mitochondrial oxygen consumption (Fig. 3A). We noted a tendency to inhibition of OXPHOS respiration in venetoclax-sensitive MOLM-13 and MV4-11 cells when combination treatment was used (Fig. 3B). In comparison, OXPHOS levels in venetoclax-resistant THP-1 cells were unaffected in the presence of single compound or combination treatments, confirming that the inhibition we observed is not a general effect of drug treatment (Fig. 3B). In addition, there was no conclusive OCR data obtained for KG-1 cell line.

Combination of IMT and venetoclax dramatically decreases OXPHOS (in a glycolysis-independent manner) in sensitive cell lines.

(A) Representative graph of an OXPHOS measurement using mitostress compounds: oligomycin (OMY), carbonyl cyanide m-chlorophenyl hydrazone (CCCP), and antimycin (AMA). (B) AML cell lines (MV4-11, MOLM-13, and THP-1) were treated with either IMT (500 nM), venetoclax (5 nM), or a combination of both for four days. DMSO was added to control wells. Oxygen consumption rate (OCR) was measured with an Oroboros O2K instrument at physiological glucose levels (5 mM) using mitostress compounds. (C) AML cell lines were cultured in growth medium supplemented with either 5 mM glucose or galactose, and cell viability tests (ATP measurements) were performed after 24 and 48 h. All results are from at least three independent experiments. Error bars show ± SEM. Statistical tests on all comparisons in (C) showed non-significant differences.

In normal hematopoietic cells, glycolysis is upregulated when OXPHOS is blocked (17). In contrast to normal hematopoietic cells, high energy turnover AML cells rely predominantly on OXPHOS due to poorly regulated metabolic adaptation. To verify the reliance on OXPHOS, AML cells were grown in growth medium supplemented with either glucose or galactose (Fig. 3C). Two ATP molecules are produced when cells use glucose for glycolysis, but if galactose is used instead, no ATP is generated (18). We observed no significant decrease in the level of ATP production (Fig. 3C) nor in cell counts (Fig. S3A) in the absence of glucose, indicating that glycolysis is not a major source of energy in AML cells. To test whether AML cells use glycolysis under treatment conditions, OCR was measured in the presence or absence of a high concentration of glucose (Fig. S3B). OCR measured for basal respiration was not significantly different with excess amounts of glucose, again indicating the inability of AML cells to switch to glycolysis when OXPHOS is disrupted.

To assess whether AML cells could rely on glycolysis for ATP generation when OxPhos was inhibited by IMT, we cultured AML cells in medium with galactose instead of glucose, a condition that limits glycolysis-dependent ATP production. Even in these glycolytic conditions, we observed minimal ATP production and cell viability, suggesting that AML cells are poorly adapted to switch to glycolysis when mitochondrial function is impaired. This finding underscores the high dependence of AML cells on OxPhos for energy production and supports the proposed mechanism of IMT efficacy, as it appears to exploit the inability of AML cells to compensate through glycolysis when mitochondrial transcription is inhibited.

Combination of IMT and venetoclax synergistically delays leukemia initiation in cell-derived leukemia mouse models

Following the confirmed synergy between IMT and venetoclax in AML cell lines, we proceeded to test the dual inhibition of mitochondrial transcription and BCL-2 in an AML cell-derived xenograft (CDX) transplantation mouse model. First, the pharmacokinetic profile of IMT was determined in plasma, liver, and bone marrow (BM) of mice, 1, 8, and 24 h post-administration (Fig. S4A). As described above (Fig. S1), the IMT mouse PK profile was superior to IMT1B. IMT was detected in all tissues at all time points when administered alone. To determine the concentration of IMT and venetoclax when administered in combination, plasma samples were then collected 8–11 h post-administration of single or combination formulations and the concentrations of IMT and/or venetoclax in samples were measured (Fig. S4B). Values obtained were all within the effective range, as combination treatment did not affect the individual compound concentrations (14). For xenograft experiments, NBSGW mice were intravenously transplanted with MV4-11 AML cells (Fig. 4A). Two weeks after transplantation, engraftment of AML cells was confirmed (evaluated by gating on hCD45+ population by flow cytometry in peripheral blood). Treatment then began with vehicle, 50 mg/kg IMT, 100 mg/kg venetoclax, or a combination of IMT (50 mg/kg) and venetoclax (100 mg/kg) (Fig. 4A), dosed as such based on previously published protocols (14, 18). The efficacy of treatments was determined by measuring the proportion of transplanted AML cells in peripheral blood and by survival analysis (Fig. 4B). In all treatment groups, the hCD45+ population dropped immediately after treatment, and vehicle- and venetoclax-treated mice then exhibited a rapid rebound. In contrast, in IMT-treated mice the increase in leukemic cell proportion was delayed until eight weeks post-transplantation, and in the combination-treated group for to up to 20 weeks (Fig. 4B). These observations correlate with results from survival analysis (Fig. 4C). IMT treatment of MV4-11 transplanted mice significantly delayed the development of leukemia, inducing a 37% increase in median overall survival (57.5 → 79 days). Development of leukemia was further delayed by combination treatment to 131% (57.5 → 133 days), and the group contained two mice that did not show any disease symptoms after 200 days (post-transplantation) when they had to be euthanized according to ethical permissions. Moreover, leukemic cells were not detectable in the BM of surviving mice at the time of termination. All other mice showed symptoms of leukemia and of these, all but one combination-treated mouse had an increased percentage of leukemic cells in BM (Fig. 4D). These results confirm that treatment with a combination of IMT and venetoclax significantly prolongs leukemia progression in mice. Moreover, the general condition of mice such as behavior, fur, mobility and weight, assessed before (week 2) and after treatment (week 5), were not altered in the IMT-, venetoclax-, or combination-treatment groups when compared to the vehicle-treated group (weights shown in Fig. S5B), suggesting that the treatments were well tolerated. As with single IMT therapy (14), treatment with combination therapy did not display any evidence of organ toxicity or drug-associated death in the mice (Fig. S5C, D).

Combination of IMT and venetoclax significantly prolongs the survival of mice with leukemia.

(A) Outline of CDX efficacy experiment: NBSGW mice were transplanted with MV4-11 human leukemia cells, then treated with either vehicle, IMT, venetoclax, or a combination of IMT and venetoclax (n=10 per treatment group). Peripheral blood was taken before treatment (week 2) and after treatment at regular intervals until disease progression was observed. (B) Analysis of proportion of human CD45+ cells in mice peripheral blood in all treatment groups at indicated time points, determined by flow cytometry. (C) Kaplan-Meier analysis of survival. Statistical significance was calculated using log-rank (Mantle-Cox) test. ****p < 0.0001, Error bars represent ± SEM. (D) Analysis of proportion of human CD45+ cells in the bone marrow of all treatment groups (four mice per group compared to two surviving mice) at time of euthanization.

IMT and venetoclax synergistically decrease the tumor burden and increase the life span in an AML-PDX mouse model resistant to venetoclax

The CDX mouse model demonstrated IMT could sensitize AML cell lines to BCL-2 inhibitors in vivo. We next analyzed the effects of combination treatment on a more refined in vivo model by generating xenografts using AML patient-derived cells (AML PDX). We first tested an AML-PDX mouse model with resistance to venetoclax (AML sample harboring TP53 mutation). Two weeks post-transplantation with AML PDX cells, NSGS mice were orally treated with IMT, venetoclax, or a combination of the two for 21 days (Fig. 5A). The efficacy of the treatments was determined based on the proportion of human AML cells (determined by hCD45+ population). Six out of 10 mice in each group were euthanized immediately after the last compound administration for end-of-treatment analysis. We found that the combination of IMT and venetoclax resulted in significant inhibition of tumor burden in blood, BM, and spleen of treated mice (Fig. 5B). By examining histology sections of BM and spleen, we found that tissues from combination-treated animals contained less engraftment of leukemic blasts than single- and vehicle-treated mice (Fig. 5C). Accordingly, the frequencies of hCD45+ cells in the BM and spleens of the combination-treated group were largely reduced in comparison to other groups (Fig. 5D), and engrafted tumor cells in combination-treated PDX mice showed a lower proliferating percentage in comparison to single or vehicle treated groups (shown by KI67+/hCD45+ immunohistochemistry staining; Fig. 5E). This demonstrates a synergy between IMT and venetoclax in reducing tumor burden.

Synergistic effect of (IMT and venetoclax) combination treatment on tumor burden and survival of AML-PDX transplanted mice.

(A) Experimental outline: NSGS mice were transplanted with primary AML splenocytes from PDX mice and were then treated with either vehicle, IMT, venetoclax, or a combination of IMT and venetoclax (n=10 per treatment group). After the final dosing, three or four mice were euthanized (end-of-treatment group) and the rest (n=6) were maintained for survival analysis. (B) Percentage of viable human CD45-positive (hCD45+) cells in peripheral blood, BM, and spleen of treated mice determined by flow cytometry. (C) Immunohistochemistry analysis of mouse BM and spleen sections stained with anti-human CD45 (brown) and counterstained with DAB (blue). (D) Quantitative analysis of hCD45+ cells on immunohistochemistry sections of BM and spleen sections of treated mice. (E) Proliferating tumor cells (CD45) are shown by quantification of Ki67 positive cells in immunohistochemistry-stained BM sections at end-of-treatment (results from at least three technical replicates and two biological replicates in each treatment group).

To further analyze the cellular effects of the IMT-venetoclax combination treatment, we analyzed transcriptomes of purified viable AML blasts isolated from PDX models at end of treatment (Fig. 6A). We confirmed that combination treatment had a more pronounced effect on gene transcription as compared to single treatments. In agreement with its more pronounced effect on tumor burden, combination treatment suppressed expression of several genes involved in leukemogenesis, including MX1, KCTD12, ITGA and RGL1 (Table S2) (19). Functional annotation and enrichment analysis were performed using the DAVID bioinformatics resources (Database for Annotation, Visualization, and Integrated Discovery) to identify significantly enriched gene ontology (GO) terms and pathways associated with the gene set. The gene list in Table S2 points to a potentially integrated function of the gene set in 1-Innate immune involvement, 2-Mitochondrial Function and Immune Signaling and 3-Signaling and Trafficking. The presence of annotations related to innate immune response, extracellular region, and transmembrane receptors suggests that the genes may be involved in the initial immune defense, recognizing pathogens, and activating immune cells. The indication of mitochondrial function and immune signaling approves that mitochondria are known not only for energy production but also for their emerging role in immune signaling, such as by producing reactive oxygen species (ROS) in response to infection. For signaling and trafficking, the presence of terms like “GTP binding,” “endosome,” and “calcium” highlights intracellular signaling, possibly in relation to immune activation or cellular responses The presence of terms like “GTP binding,” “endosome,” and “calcium” highlights intracellular signaling, possibly in relation to immune activation or cellular responses. We also analyzed survival by monitoring the mice for disease progression and morbidity. Peripheral blood engraftments were analyzed at regular intervals until the appearance of disease symptoms (Fig. 6B and Fig. S6A). The leukemia burden immediately prior to euthanization demonstrated that the cytoreductive effect of the combination treatment persists (Fig. 6B). However, blood count (determined by RBCs and HGB levels) did not vary significantly between experimental groups, nor did the magnitude of splenomegaly or liver weight vary at the time of euthanization (Fig. S6B–D). Interestingly, mice treated with a combination of IMT and venetoclax succumbed to death significantly later than other treatment groups, with a median survival of 112 days vs 96, 99, and 102 days for vehicle-, IMT-, and venetoclax-treated groups, respectively (Fig. 6C). Combination treatment thus increases the life span of venetoclax-resistant AML mouse models, demonstrating that IMT can sensitize AML cells towards other therapies.

Synergistic effect of (IMT and venetoclax) combination treatment on gene transcription and survival of AML-PDX transplanted mice.

The experimental outline was shown in figure 5A, but n=10 per treatment group. After the final dosing, three or four mice were euthanized (end-of-treatment group) and the rest (n=6) were maintained for survival analysis. (A) Gene expression profiling by RNA-seq of purified AML blasts isolated from PDX models treated with venetoclax, IMT, or a combination thereof. Indicated expression changes are relative to a vehicle treated group. Significantly enriched genes are indicated in red (q < 0.05). (B) Peripheral blood engraftment analysis of survival group. (C) Survival analysis of treated mice shown by Kaplan-Meier analysis (n=6/treatment group). *P < 0.05, **P < 0.01, and ***p < 0.001.

IMT does not affect tumor burden or life span in an AML-PDX mouse model sensitive to venetoclax

For comparison, we also monitored effects of combination treatment on a PDX model based on AML cells sensitive to venetoclax treatment (AML cells harboring FLT3-ITD, NMP1 and DNMT3A). As expected, we observed that venetoclax treatment significantly reduced the tumor burden in the blood and BM of mice. Combination treatment with IMT did not add to this effect (Fig. 7A). By examining histology sections, we found that both BM and spleen tissues from combination- and venetoclax-treated animals contained reduced proportion of leukemic blasts compared to IMT- and vehicle-treated mice (Fig. 7B). White blood counts (WBCs), red blood counts (RBCs), and hemoglobin (HGB) levels were not significantly different across experimental groups (Fig. S7A). The total body and liver weight of PDX mice were very similar in all groups (Fig. S7B, D), but both venetoclax and combination treatments reduced splenomegaly (Fig. S7C).

Leukemia progression induced by primary AML cells is inhibited by venetoclax treatment.

As in Fig. 5A, efficacy experiment outline: NSGS mice were transplanted with primary AML splenocytes from PDX mice and were then treated with either vehicle, IMT, venetoclax, or a combination of IMT and venetoclax (n=12 per treatment group). 8–12 h after the last dosing, half of the mice (n=6/treatment group) were euthanized (end-of-treatment group) and the other half were maintained for survival analysis. (A) Percentage of viable human CD45-positive (hCD45+) cells in peripheral blood and bone marrow of treated mice (n=6 for each treatment group), determined by flow cytometry. (B) Immunohistochemistry analysis of hCD45+ cells on bone marrow and spleen tissue sections of treated mice (images are representative from at least two mice per group). Scale bars represent 100 μM. (C) Peripheral blood engraftment analysis of survival group. (D) Mice treated with single or a combination of compounds showed prolonged survival in comparison to the vehicle-treated group. Kaplan-Meier analysis shows the relative survival curve. *P < 0.05, **P < 0.01, and ***p < 0.001.

The remaining mice were used for survival analysis, which involved regular blood count and chimerism analysis until the appearance of disease symptoms (Fig. 7C). Both venetoclax and IMT caused a modest increase in life span, but combination treatment did not offer an additional effect (Fig. 7D).

Discussion

In recent years, the genetic complexity and heterogeneity of leukemia cell populations has become apparent (20, 21) and developing effective strategies to target human leukemia initiating cells has thus proven to be a challenging task. Currently, there is considerable interest in therapies, such as venetoclax, that can restore normal apoptosis of cancer cells by directly interacting with the BCL-2 family of proteins. These are effective for some leukemias, but many patients eventually become refractory to treatment and ultimately succumb to their disease (22, 23). It has become increasingly important to determine which biological properties are relatively consistent in leukemic cell populations irrespective of genetic makeup (tumor population). Recent observations indicate that several cancer cell types, including AML, are reliant on OXPHOS for metabolism (24), making inhibition of mitochondrial function a promising therapeutic strategy for AML (12, 25–28). Therapy-resistant leukemic cells are more reliant on oxidative respiration than glycolysis for energy generation, suggesting that the OXPHOS dependency and lack of normal glycolytic activity in AML cells may be characteristic of cancers (29). Understanding resistance mechanisms will allow for the design of strategies to counteract drug resistance or impede resistance from developing. In AMLs with aberrant expression of BCL-2, venetoclax can reduce OXPHOS (9) and trigger cell death. However, venetoclax resistance can occur in patients treated with the BCL-2 inhibitor and in some subtypes of AML patients with poor prognoses (30).

Previous work has indicated that AML cells display increased mitochondrial biogenesis and inhibition of mitochondrial translation can be used as a strategy for human AML (18, 27). In our work, we target another process involved in mitochondrial gene expression, i.e., transcription. IMT treatment targets POLRMT, which has two main functions in mitochondria: it transcribes mtDNA encoded genes and provides the primers required to initiate mtDNA replication. Inhibition of POLRMT can thus deliver a double punch, both reducing mtDNA copy number and inhibit OXPHOS biogenesis. As demonstrated here inhibition of POLRMT can overcome venetoclax therapy resistance in AML. These observations are in line with recent reports of the importance of mitochondrial function in leukemia cell survival (31, 32). By developing a clinically relevant and well-tolerated approach to deliver IMTs in immunodeficient mice transplanted with AML cells, we were able to reduce the tumor burden in all (cell- and patient-derived) xenografted mice treated with a combination of IMT and venetoclax. This demonstrates that reducing POLRMT activity results in successful targeting of leukemic cells. Our in vitro investigations confirmed that the combination of IMT and venetoclax is necessary to significantly decrease ATP levels and consequently induce apoptosis. In addition, efficacy studies of AML PDX models showed that single compounds are not sufficient to eradicate AML cells, but that the combination regimen reduced AML tumor burden and prolonged survival in all treated mice. These data indicate the advantage of both downregulating POLRMT activity in combination with increasing apoptosis and highlights the importance of rational combination therapies (33, 34). IMT offers potential advantages over other OXPHOS-targeting therapies due to its lower toxicity, as observed in preclinical models. While other strategies, such as mitochondrial translation inhibition or electron transport chain inhibitors, effectively target OXPHOS in AML, they may come with higher toxicity risks. IMT’s selective targeting of mitochondrial transcription appears to minimize adverse effects, making it a promising candidate for use in combination regimens with BCL-2 inhibitors like venetoclax. This approach could improve therapeutic outcomes while maintaining tolerability, addressing a significant challenge in AML treatment.

In our study, we observed that the combination of IMT and venetoclax did not show synergistic effects in TP53 mutant AML cells, particularly in the THP-1 cell line. Although statistical analyses indicated potential interaction, the biological impact was minimal in TP53 mutant cells, contrasting with the robust synergy observed in TP53 wild-type cells like MV4-11 and MOLM-13. This suggests that TP53 mutation status may influence the efficacy of this combination therapy, highlighting a potential limitation in targeting mitochondrial transcription to potentiate BCL-2 inhibition in TP53 mutant AML context. These findings underscore the need for broad mutational profiling when designing combination therapies for AML.

In summary, we have introduced a new therapeutic strategy to target therapy-resistant leukemia with OXPHOS-dependent properties. Despite several limitations with the design of pre-clinical cancer models and the intensity and duration of treatments, we demonstrate here that the combined inhibition of BCL-2 and mitochondrial transcription can target therapy-resistant AMLs. This combination treatment succeeds as IMT primes cell vulnerability to apoptosis induction. Consequently, our findings may be of value for the future development of better clinical strategies for the treatment of AML.

Materials and methods

Cell lines and culture conditions

All cell lines (MV4-11, MOLM13, THP-1, KG-1, OCI-AML3, and OCI-AML2) were purchased from DSMZ (Braunschweig, Germany) and cultured in RPMI-1640 (Sigma Aldrich) supplemented with 10–20% heat inactivated fetal bovine serum (Gibco) and 1% penicillin/streptomycin (Invitrogen). Cells were maintained in a humidified incubator at 37 °C and 5% CO2. RPMI-1640 without glucose was used for in vitro treatment experiments and supplemented by physiological concentrations (5 mM) of glucose (D-Glucose, Sigma) or galactose (D-galactose, Sigma).

Generation of in vivo AML mouse model

Mice were purchased from Jackson laboratories (USA) and maintained in a pathogen-free facility at the University of Gothenburg. All animal experiments were conducted in accordance with guidelines approved by the research animal ethics committee at the University of Gothenburg. For AML cell line xenograft (CDX) experiments, 6–8-week-old, male NBSGW mice (NOD.Cg-KitW-41J Tyr+ Prkdcscid Il2rgtm1Wjl/ThomJ, Jackson laboratories) were used. Mice were transplanted intravenously with ∼1 × 106 MV4-11 cells. For patient-derived xenograft (PDX) experiments, 6–8-week-old, male NSGS mice (NOD.Cg-Prkdcscid Il2rgtm1Wjl Tg (CMV-IL3,CSF2,KITLG)1Eav/MloySzJ, Jackson laboratories) were sub-lethally irradiated with 1 Gy radiation. AML PDX splenocytes were purchased from Jackson laboratories (id: J000106134*, male, 58 years old, AML, M4, positive for FLT3-ITD, DNMT3A, and NPM1mutaions; and id: 000106566*, male, 48 years old, AML no classification specified, positive for TP53 mutation). To obtain adequate number of PDX cells for the scale of the efficacy experiments, the splenocytes were initially expanded in vivo. Mice were transplanted intravenously with ∼1 × 106 splenocytes from AML PDX mice containing more than 80% patient AML cells.

Human leukemia cell frequency assessment by flow cytometry

The percentage of human leukemic cells (MV4-11 cell lines and AML PDX cells) in mouse peripheral blood, BM, and spleen was determined by flow cytometry. Mice were bled at indicated time points after transplantation. Prior to staining, peripheral red blood cells were lysed using ammonium chloride for 8–10 min at room temperature. Cells were stained with antibodies, washed, and submitted for flow-cytometric analysis using a FACS Aria (BD Biosciences, USA) or CytoFlex instrument (Beckman Coulter, California, USA). At least 10,000 events were acquired, and the percentage was calculated relative to gated viable mononuclear cells (MNCs). To discriminate between human and mouse cells, cells were stained with antibodies against human CD45 (clone HI30, BD Biosciences) and mouse CD45 (clone 2D1, BD Biosciences). Engraftment was deemed positive if human CD45+ cells represented >0.1% of the viable MNCs. 7AAD (BD Biosciences) served as viability marker.

Preparation of compounds for in vivo drug administration

IMT was optimized based on the structure activity and property relationship of the coumarin mtRNAPol inhibitor class by LDC (Dortmund, Germany). Venetoclax was obtained from MedChemExpress (Hölzel Diagnostika Handels GmbH, Germany). Following the development of a suitable vehicle for the formulation of IMT and venetoclax, as well as the combination of both compounds, the test articles were formulated using 0.5% aqueous methylcellulose solution (Sigma-Aldrich, Germany). After randomization, and two weeks post-transplantation or with confirmation of engraftment, mice were treated with vehicle (0.5% methylcellulose), 50 mg/kg IMT, 100 mg/kg venetoclax, or a combination (100 mg/kg venetoclax + 50 mg/kg IMT) by oral administration daily for 21 days.

Cell proliferation assay

Cells were seeded out ca 20 h prior to treatment. For dose response experiments, cells were treated with a wide range of concentrations of single compounds. Cell proliferation assays were performed every 24 h from 72–168 h, where cell viability/ATP production was determined using CellTiter-Glo (Promega).

Apoptosis assay and Mitochondrial potential assay

Apoptosis was evaluated by Annexin V and PI staining and mitochondrial membrane potential was evaluated by lipophilic cationic dye JC-1. Cells were treated with IC50 concentrations of IMT, venetoclax, or combination of both, analyzed using an advanced image cytometer (Nucleocounter NucleoCounter’s accompanying software NucleoView).

High-resolution respirometry

AML cell lines were treated with DMSO, IMT, venetoclax, or combination of IMT and venetoclax for 3–4 days. The mitochondrial respiratory capacity (oxygen consumption rate) was measured using high resolution respirometry (O2K, Oroboros, Innsbruck, Austria). A substrate-uncoupler-inhibitor titration (SUIT) protocol was used to measure specific features of mitochondrial respiration. Respiratory capacities were tested in a sequence of coupling states: basal respiration; LEAK respiration using 5 mM oligomycin (Sigma); and non-coupled electron transfer state (ET state) where the optimum carbonyl cyanide 3-chlorophenylhydrazone (CCCP, Sigma) concentration for maximum flux was determined by titration (max. 3 μM CCCP). Mitochondrial respiration was inhibited using 2.5 μM antimycin (Sigma). Where stated, excess glucose (D-Glucose, Merck) was added at a final concentration of 20 mM. Instrumental and oxygen background fluxes were calibrated as a function of oxygen concentration and subtracted from the total volume-specific oxygen flux. Calibration and data analysis was performed using Datlab v.6.1 software (Oroboros Instruments).

Pharmacokinetics analysis

IMT and venetoclax were extracted from plasma and tissues by protein precipitation using acetonitrile. Tissue samples were homogenized using double the (sample) volume of PBS, and bone marrow samples were suspended in 40 µL of PBS prior to extraction with acetonitrile. Samples were analyzed by liquid chromatography tandem-mass spectrometry using a Prominence UFLC system (Shimadzu) coupled to a Qtrap 5500 instrument (ABSciex). Test articles were separated on a C18 column using a gradient elution with an acetonitrile/water mixture containing 0.1% formic acid as the mobile phase. Chromatographic conditions and mass spectrometry parameters were optimized for each test article prior to sample analysis. Concentrations of each test article were calculated by means of a standard curve.

mRNA extraction and transcriptome sequencing of AML PDX cells

For RNAseq experiment BM samples were FACS sorted using BD FACS Aria fusion (BD Biosciences). Human AML cells were isolated using human CD45 antibody. 7-AAD (BD Biosciences) was used for dead cell exclusion. RNA was extracted using RNeasy Micro kit (Qiagen). RNAseq library preparation and RNA-sequencing was done using the SMARTer Total Stranded RNA pico kit. RNA library was prepared at National Genomics Infrastructure (NGI) to be sequenced on the NovaSeq S6000 on 0.5 lanes of one S4 flowcell.

Bioinformatics analysis

Analysis was performed in R (www.r-project.org, accessed on 10 May 2022) using DESeq2, clusterProfiler and ReactomePA packages. Prefiltering to keep genes that have at least 10 reads total was performed before differential expression analysis was done with DESeq2. Visualization of differentially expressed genes was done with EnhancedVolcano to create volcano plots and gridExtra for creating tables. The p-value cutoff was set to 0.1 to allow for more genes to be found in EP (End-of-treatment) group comparisons and the Res/Sen (Resistant/Sensitive) group comparisons. Upregulated and downregulated genes were plotted and colored according to Fold Change (FC) greater than 1.5 or -1.5 and the p-value cutoff 0.1.

Functional annotation and enrichment analysis were performed using the DAVID bioinformatics resources (Database for Annotation, Visualization, and Integrated Discovery) to identify significantly enriched gene ontology (GO) terms and pathways associated with the gene set (35).

Statistical Analysis

Unless otherwise noted, data were summarized using mean ± standard error of the mean (± SEM). Per-group sample sizes presented in figures and results are reported from ≥3 separate experiments, unless stated otherwise. Survival distributions were estimated using the Kaplan– Meier method, and group comparisons were conducted using the log-rank test. Data were analyzed using GraphPad Prism 9.0. Combination analysis was summarized using the Bliss-independent model (16). The formula Yab,P = Ya +Yb − YaYb was used where drug A at dose a inhibits Ya percent of cell growth, and drug B at dose b inhibits Yb percent of cell growth. The predicted inhibitory effect of combined percentage was estimated by Yab,P. The observed combined percentage inhibition Yab,O was then compared with Yab,P, and if Yab,O > Yab,P, the combination treatment was thought to be more efficacious (synergy). If Yab,O = Yab,P the combination treatment was thought to be additive.